Towards a standardized evaluation of imputation methodology

### 2022-09-16 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
changes

For some data analyses a single imputation may suffice, while for other data analyses multiple imputations are needed. For example, in the case of inferential analyses, such multiple imputations \citep{rubi76} have proven to be a valuable technique for obtaining valid inferences on incomplete datasets. The resulting multiple inferences on multiple completed data sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. When generating predicted values is the goal of the data analysis, a single imputation may already generate sufficient \citep{nijm20} predicted values.

@@

The purpose of this paper is threefold: First, we raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as reviewers for statistical journals. Second, we provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground, but is in no way intended as an absolute solution. This identifies the third purpose of our paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties.

IK HEB EEN BEETJE MOEITE MET ONDERSTAANDE PARAGRAAF: HET GAAT OM DE REF VAN TIM EN DE PREREG. DIE KUNNEN WE OOK 2 ALINEAS EERDER INFIETSEN. IK HEB EEN POGING GEDAAN.
Standardizing the evaluation of imputation methodology requires simulators adopt common aspects in their simulation and evaluation workflows. Table \ref{table:check} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed structure for planning and reporting simulation studies and \citet{pawe22}'s advice to pre-register simulation protocols. We henceforth highlight and add some recommendations for the evaluation of imputation methodology. 
% [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

A well-designed simulation does not guarantee fair evaluations. Still, we propose to consider the following simulation design choices --GV WHICH CHOICES? Ik snap deze eerste zin niet helemaal--. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citet{pete14}. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions) [TODO: add ref dance of the mechanisms!]. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

To evaluate the ability of an imputation routine to handle missingness, a form of ground  truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled (with replacement) GV IS DIT NODIG? from a sufficiently large observed set, such as official registers. The simulator should choose their data-generating mechanism in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

@@

Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) GV TEVEEL HAAKJES comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s), and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be very challenging. Most real-world data contains at least some missing entries, for which the true underlying missing data model is--by definition--unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It may seem like an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown missing data model. Only sampling complete cases from the data set may thus result in samples that fail to capture all relevant real-world conditions, which in turn refutes the main reason for using design-based simulation. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Just to be clear: leaving the missingness as-is and inducing additional missing values to impute is no option here, because we would not have a real and unbiased comparative truth.

@@

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness). Simulators should therefore be explicit in their description of the missing data generating model fitting the aim of their simulations.

@@

[TODO: add that focusing on a single mechanism (even if the mechanism suits the study aim) is wasteful?] Truly random missingness across all data entries (`missing completely at random' or MCAR under Rubin's definition) may be considered a necessary simulation condition for the evaluation of imputation procedures, since the statistical properties of the observed data given the missing data are known. Any imputation routine that cannot at least mimic the performance of the observed data inference should be deemed inefficient in the scope of the simulation. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not universally sound. Sadly, the straightforward case of MCAR is often neglected from simulation studies, while it offers an informative reference condition in many simulation setups.

[TODO: maybe add definitions of missingness mechanisms back in???] A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}. Inducing a MAR mechanism presents the simulator with a choice, namely the type or functional form of the missingness model. Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model \citep{scho18}. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single functional form of the missingness-generating model.

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.

>>>>

For some data analyses a single imputation may suffice, while for other data analyses multiple imputations are needed. For example, in the case of inferential analyses, such multiple imputations \citep{rubi76} have proven to be a valuable technique for obtaining valid inferences on incomplete datasets. The resulting multiple inferences on multiple completed data sets can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. When generating predicted values is the goal of the data analysis, a single imputation may already generate sufficient predicted values \citep{nijm20}.

@@

The purpose of this paper is threefold: First, we raise some concerns with respect to evaluating imputation methodology. These concerns stem from careful consideration with fellow `imputers' and from encounters as reviewers for statistical journals. Second, we provide imputation methodologists with a suggested course of action when using simulation studies to evaluate imputation techniques for missing data problems. This suggested approach should identify common ground, but is in no way intended as an absolute solution. This identifies the third purpose of our paper: discussion. We hope to elicit critical thinking regarding the problems at hand. We are all convinced that our methodology has some merit. But for sake of progress, it would be much more advantageous if the aim of our evaluations would go beyond \textit{proving the point} and would legitimately consider the statistical properties. 
% [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

A well-designed simulation does not guarantee fair evaluations. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citet{pete14}. The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions) [TODO: add ref dance of the mechanisms!]. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

To evaluate the ability of an imputation routine to handle missingness, a form of ground  truth has to be established. Those who perform simulation studies are in the luxury position to establish the truth beforehand by choosing a data-generating mechanism. Data-generating mechanisms define how a complete dataset is obtained at the start of each simulation repetition. There are two general approaches to generating complete data: (i) model-based simulation, in which data are drawn from a known statistical model or probability distribution, such as the multivariate normal distribution; and (ii) design-based simulation, where data are sampled from a sufficiently large observed set, such as official registers. The simulator should choose their data-generating mechanism in line with the study scope. Model-based data-generating mechanisms have advantages in flexibility and precision, since data are generated from a known statistical model and the true theoretical parameters can be derived. A design-based approach is often used in situations where a probability distribution is not available, or where real-life data structures are of interest.

@@

Another threat to the generalizability of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) GV TEVEEL HAAKJES comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the measurement level, and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be challenging. Most real-world data contains at least some missing entries, for which the true underlying missing data model is--by definition--unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It may seem like an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown missing data model. Only sampling complete cases from the data set may thus result in samples that fail to capture all relevant real-world conditions, which in turn refutes the main reason for using design-based simulation. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Just to be clear: leaving the missingness as-is and inducing additional missing values to impute is no option here, because we would not have a real and unbiased comparative truth.

@@

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term `missingness proportion' restricted to the variable-by-variable interpretation. The distinction between the two concepts, however, is not always clear in the literature, and convoluting the terms may lead to incorrect generalizations because they rarely mean the same (e.g., a proportion of 50\% incomplete cases in bivariate data could translate to a missingness proportion of 25\% in both variables, or one completely observed variable and one with 50\% missingness). Simulators should therefore be explicit in their description of the missing data generating model and should make sure that the model fits the aim of their simulations.

@@

Truly random missingness across all data entries (`missing completely at random' or MCAR under Rubin's definition) may be considered a necessary simulation condition for the evaluation of imputation procedures, since the statistical properties of the observed data given the missing data are known. Any imputation routine that cannot at least mimic the performance of the observed data inference should be deemed inefficient in the scope of the simulation. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not universally sound. Sadly, the straightforward case of MCAR is often neglected from simulation studies, while it offers an informative reference condition in many simulation setups.

GV STREAMLINEN [TODO: maybe add definitions of missingness mechanisms back in???] A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly}. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}. Inducing a MAR mechanism presents the simulator with a choice, namely the type or functional form of the missingness model. Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model \citep{scho18}. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single functional form of the missingness-generating model.

Although MAR missingness is often considered as a simulation condition, the problem of spurious MAR is generally overlooked. With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on observed gender to emulate a situation wherein one gender is less likely to disclose their weight). These relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the true data, claims for a method's applicability to situations where the missingness is random become less valid. The most extreme example would be when MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is generated random GV PLAATJE WEER ER IN BRENGEN!. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than imputation \citep[see e.g.][\S 2.7]{buur18}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is pointless.
